hello
  indented
line3---
title: "Examining the Impact of Race and Gender on Emergency Room Treatment"
subtitle: "STA 230 Final Project: analysis code"
author: "Zachary Serby"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Packages

```{r packages}
library(tidyverse)     # dplyr, ggplot2, stringr, tibble, etc.
library(readxl)
library(tidytext)
library(broom)
library(nnet)          # multinomial logistic regression
library(rpart)         # decision trees
library(rpart.plot)
library(randomForest)
library(pROC)          # ROC curves / AUC
library(caret)
```

## Load and clean the NEISS data

The yearly NEISS extracts (`NEISS_2013.XLSX` ... `NEISS_2022.XLSX`) go in `data/raw/`,
and the NEISS format key (`NEISS_FMT.XLSX`) goes in `data/`. See `data/README.md`.

```{r load-data}
## List all yearly NEISS files
file_list <- list.files(path = "../data/raw", pattern = "\\.XLSX$", full.names = TRUE)

## Initialize an empty list to store data frames
dataframe_list <- list()

## Loop through each file and read data
for (file_name in file_list) {
  # Read data from Excel file
  data <- read_excel(file_name)

  ## Extract year from file name and convert to integer
  Year <- as.integer(sub("^NEISS_(\\d{4})\\.XLSX", "\\1", basename(file_name)))

  ## Add year column to the data
  data$year <- Year

  ## Append the data frame to the list
  dataframe_list <- append(dataframe_list, list(data))
}

## Combine all data frames into a single data frame
combine_data <- bind_rows(dataframe_list)

## Drop unused columns (sample weights, PSU/stratum, mostly-NA fields) and
## keep records with Sex = 1 (male) or 2 (female)
cleaner_data <- combine_data[ -c(6, 7, 10:13, 15:21, 23:25) ] %>%
    filter(!Sex %in% c(0, 3))

cleaner_data$Disposition <- as.factor(cleaner_data$Disposition)
cleaner_data$Race <- as.factor(cleaner_data$Race)
cleaner_data$Sex <- as.factor(cleaner_data$Sex)

## Change the reference level of the categorical variable to "1 = White"
cleaner_data$Race <- relevel(cleaner_data$Race, ref = "1")
```

```{r load-key}
## Read and tidy the NEISS format key
key_file <- "../data/NEISS_FMT.XLSX"
key_data <- read_excel(key_file)

## Rename column names
colnames(key_data) = c("format_name", "start_value", "end_value", "format_value")

## Remove variables not used in this analysis
new_key <- key_data %>%
  filter(!format_name %in% c("PROD", "LOC", "HISP", "FIRE"))
```

## Tests of association

Do all sexes and races have an equal chance of each treatment (disposition)?

Null hypothesis: the probability of getting any treatment is the same regardless of sex.
$H_0: P(\text{Treatment} \mid \text{Sex}) = P(\text{Treatment})$

```{r chisq-sex}
## Chi-squared test of association
chisq.test(x = table(cleaner_data$Disposition, cleaner_data$Sex))
```

Null hypothesis: the probability of getting any treatment is the same regardless of race.
$H_0: P(\text{Treatment} \mid \text{Race}) = P(\text{Treatment})$

```{r fisher-race}
## Fisher's Exact Test, looking at the variables Race and Disposition
fisher.test(x = table(cleaner_data$Disposition, cleaner_data$Race), simulate.p.value = TRUE)
```

I obtained a p-value < 2.2e-16 for Sex and Disposition from a chi-squared test, and a
p-value of 0.0004998 for Race and Disposition from Fisher's exact test (simulated p-value).

Both p-values are much smaller than the common significance level of 0.05. Therefore, there is
sufficient evidence to reject both null hypotheses that the probability of getting any
disposition (treatment) is the same regardless of sex and of race. There is enough statistical
evidence to further investigate the association between Race/Sex and treatment.

## Comparing models: multinomial logistic regression, decision tree, random forest

```{r fit-models}
## Fit a multinomial logistic regression model
mfit <- multinom(Disposition ~ Race + Sex, data = cleaner_data, trace = FALSE)

## Fit a decision tree
my_tree <- rpart(Disposition ~ Race + Sex, data = cleaner_data)

## Fit a random forest
rf <- randomForest(Disposition ~ Race + Sex, data = cleaner_data, ntree = 300)
```

```{r accuracy}
## Function to calculate accuracy
calculate_accuracy <- function(model, data) {
  ## Get the predicted class probabilities
  fitted_probs <- predict(model, newdata = data, type = "prob")

  ## Find the index of the maximum probability for each observation
  most_likely <- apply(fitted_probs, 1, which.max)

  ## Map the indices to category labels
  pred_cats <- levels(data$Disposition)[most_likely]

  ## Calculate the fraction of correct predictions (accuracy)
  accuracy <- sum(pred_cats == data$Disposition) / nrow(data)

  return(accuracy)
}

## Calculate accuracy for each model
accuracy_mfit <- calculate_accuracy(mfit, cleaner_data)
accuracy_tree <- calculate_accuracy(my_tree, cleaner_data)
accuracy_rf <- calculate_accuracy(rf, cleaner_data)

## Print the accuracies
cat("Multinomial Logistic Regression Accuracy:", accuracy_mfit, "\n")
cat("Decision Tree Accuracy:", accuracy_tree, "\n")
cat("Random Forest Accuracy:", accuracy_rf, "\n")
```

Multinomial logistic regression did not outperform the classification trees on accuracy: no
splits occurred in the decision tree, and all three models have the same accuracy (0.9022879).

Because the accuracies are identical, we compare the models using the AUC for each disposition.

### ROC / AUC: multinomial logistic regression

```{r roc-multinom}
## Predicted probabilities for each class (each disposition)
predicted_probs <- predict(mfit, newdata = cleaner_data, "probs")

## Extract the true class labels
true_labels <- cleaner_data$Disposition

## One-vs-rest ROC curve for each class
roc_multiclass <- lapply(1:nlevels(true_labels), function(i) {
  roc(as.numeric(true_labels == levels(true_labels)[i]), predicted_probs[, i], quiet = TRUE)
})

## Plot the multiclass ROC curves
colors <- c("red", "green", "blue", "purple", "yellow", "black")
for (i in 1:length(roc_multiclass)) {
  plot(roc_multiclass[[i]], col = colors[i], add = i > 1)
}
legend("bottomright", legend = levels(true_labels), fill = colors)

## Calculate AUC for each class
auc_values <- sapply(roc_multiclass, function(roc_class) {
  auc(roc_class)
})
print(auc_values)
```

AUC values range from 0.544 for Disposition 1 (treated and released) to 0.699 for
Disposition 8 (died in the ER), all above 0.5. An AUC of 1 indicates perfect classification,
while an AUC of 0.5 is equivalent to random guessing.

### ROC / AUC: decision tree

```{r roc-tree}
predicted_probs <- predict(my_tree, newdata = cleaner_data, "prob")
true_labels <- cleaner_data$Disposition

roc_multiclass <- lapply(1:nlevels(true_labels), function(i) {
  roc(as.numeric(true_labels == levels(true_labels)[i]), predicted_probs[, i], quiet = TRUE)
})

for (i in 1:length(roc_multiclass)) {
  plot(roc_multiclass[[i]], col = colors[i], add = i > 1)
}
legend("bottomright", legend = levels(true_labels), fill = colors)

auc_values <- sapply(roc_multiclass, function(roc_class) {
  auc(roc_class)
})
print(auc_values)
```

The decision tree has an AUC of 0.5 for every disposition, lower than the multinomial
logistic regression model.

### ROC / AUC: random forest

```{r roc-rf}
predicted_probs <- predict(rf, newdata = cleaner_data, "prob")
true_labels <- cleaner_data$Disposition

roc_multiclass <- lapply(1:nlevels(true_labels), function(i) {
  roc(as.numeric(true_labels == levels(true_labels)[i]), predicted_probs[, i], quiet = TRUE)
})

for (i in 1:length(roc_multiclass)) {
  plot(roc_multiclass[[i]], col = colors[i], add = i > 1)
}
legend("bottomright", legend = levels(true_labels), fill = colors)

auc_values <- sapply(roc_multiclass, function(roc_class) {
  auc(roc_class)
})
print(auc_values)
```

The random forest also has an AUC of 0.5 for every disposition, like the decision tree.
We therefore use the multinomial logistic regression model as the one that best represents
our data.

## Multinomial logistic regression

```{r multinom-summary}
## Summary of the data
summary(cleaner_data)

## Fit a multinomial logistic regression model
mfit <- multinom(Disposition ~ Race + Sex, data = cleaner_data, trace = FALSE)

## Coefficients (log relative risk ratios vs. Disposition 1, White, Male)
coef(mfit)

## Exponentiated coefficients for Race + Sex (relative risk ratios)
exp(coef(mfit)[, 2:8])
```

Note: the key lists 8 dispositions, but "no injury" (0) and "unknown" (9) have zero patients.

There's a decrease in the likelihood of being treated and transferred (Disposition 2) for Black
patients (Race 2) relative to White patients (Race 1, the reference level), and for women
relative to men.

```{r multinom-pvalues}
## Convert model coefficients to a tidy data frame
tidy_mfit <- tidy(mfit)

## Coefficient p-values by disposition, colored by predictor
ggplot(tidy_mfit, aes(x = p.value, y = factor(y.level), color = term)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  annotate("text", x = 0.06, y = Inf, vjust = 1.5, label = "p = 0.05", color = "red", size = 3, hjust = 0) +
  labs(title = "Multinomial Logistic Regression Coefficients",
       x = "p-value",
       y = "Treatment",
       color = "Predictor Variable") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The purple marker at Disposition 8 is a p-value below 0.05, so we reject the null hypothesis that
Native Hawaiian/Pacific Islander patients are as likely as White patients to die in the ER.
This disparity warrants further investigation.

## Decision tree

```{r tree}
## No adjustment of cp or minsplit changes this tree
my_tree <- rpart(Disposition ~ Race + Sex, data = cleaner_data)
prp(my_tree, extra = 104)
```

The single node predicts TREATED/EXAMINED AND RELEASED (Disposition 1). About 90% of patients
were treated and released, 2% treated and transferred, 6% admitted, 1% held for observation,
2% left without being seen, and ~0% died in the ER.

## Random forest

```{r rf}
rf <- randomForest(Disposition ~ Race + Sex, data = cleaner_data, ntree = 300)
rf$importance   # variable importance (mean decrease in Gini)
plot(rf)
```

The error rate doesn't improve as trees are added because the forest always predicts the
dominant category (Disposition 1, treated and released). Sex has a larger Gini importance
than Race.

## Additional analysis

### Most common words in the injury narrative, by sex and race

The wording in the `Narrative` field differs by race and sex. Terminology could be one factor
behind disparities in dispositions.

```{r words-sex}
narrative_sex_data <- cleaner_data %>%
  select(Narrative, Sex)

## Split narrative into individual words
words <- str_split(narrative_sex_data$Narrative, "\\s+")

## Tidy format: one row per word, paired with narrative id and Sex
tidy_form <- tibble(word = unlist(words),
                    row = rep(1:length(words), lengths(words)),
                    Sex = rep(narrative_sex_data$Sex, lengths(words)))

## Keep words with four letters or more
tidy_form_filtered <- tidy_form %>%
  filter(str_length(word) >= 4)

## Word frequencies by Sex
freq_word <- tidy_form_filtered %>%
  group_by(Sex, word) %>%
  count(sort = TRUE)

## Top 10 words for each Sex
pop_words_sex <- freq_word %>%
  group_by(Sex) %>%
  slice_max(n = 10, order_by = n)

print(pop_words_sex)
```

```{r words-race}
narrative_race_data <- cleaner_data %>%
  select(Narrative, Race)

words <- str_split(narrative_race_data$Narrative, "\\s+")

tidy_form <- tibble(word = unlist(words),
                    row = rep(1:length(words), lengths(words)),
                    Race = rep(narrative_race_data$Race, lengths(words)))

tidy_form_filtered <- tidy_form %>%
  filter(str_length(word) >= 4)

freq_word <- tidy_form_filtered %>%
  group_by(Race, word) %>%
  count(sort = TRUE)

pop_words_race <- freq_word %>%
  group_by(Race) %>%
  slice_max(n = 10, order_by = n)

print(pop_words_race)
```

### Sex and race composition of each disposition

```{r composition}
segment_bar_data <- cleaner_data %>%
  select(Disposition, Sex, Race) %>%
  mutate(
    Disposition = case_when(
    Disposition == 1 ~ "TRE/REL",
    Disposition == 2 ~ "TRE/TRANS",
    Disposition == 4 ~ "TRE/HOSP",
    Disposition == 5 ~ "HELD",
    Disposition == 6 ~ "NOTSEEN",
    Disposition == 8 ~ "DIED",
    Disposition == 9 ~ "UNKNOWN"
    ))

## Segmented bar chart by sex
ggplot(segment_bar_data, aes(x = Disposition, fill = Sex)) +
  geom_bar(position = "fill") +
  scale_fill_discrete(labels = c("Male", "Female"))

## Segmented bar chart by race
ggplot(segment_bar_data, aes(x = Disposition, fill = Race)) +
  geom_bar(position = "fill") +
  scale_fill_discrete(labels = c("N.S.", "WHITE", "BLACK", "OTHER", "ASIAN", "AMER.IND./ALS.NAT.", "PAC.ISL"))
```

### Multinomial logistic regression without Disposition 1

Disposition 1 (treated and released) is the large majority, so we check whether Race and Sex
remain significant without it.

```{r without-treatment1}
Removed_treatment1 <- cleaner_data %>%
  filter(Disposition != 1)
Removed_treatment1$Race <- relevel(Removed_treatment1$Race, ref = "1")

mfit <- multinom(Disposition ~ Race + Sex, data = Removed_treatment1, trace = FALSE)
tidy_mfit <- tidy(mfit)

ggplot(tidy_mfit, aes(x = p.value, y = factor(y.level), color = term)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  annotate("text", x = 0.06, y = Inf, vjust = 1.5, label = "p = 0.05", color = "red", size = 3, hjust = 0) +
  labs(title = "Multinomial Logistic Regression Coefficients (Disposition 1 removed)",
       x = "p-value",
       y = "Treatment",
       color = "Predictor Variable") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

With Disposition 1 omitted, Race and Sex are less significant (including for female and Native
Hawaiian/Pacific Islander patients). This suggests much of the overall race/sex disparity is
driven by who is treated and released. The result may also reflect the change in race and sex
proportions after removing those rows.
